---
title: "2 Time Series Graphics"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(fpp3)
```

## PBS Data

The PBS dataset contains sales data on pharmaceutical products in Australia by month.

```{R}
head(PBS)
```

-   The data set shows Medicare prescription data from July 1991

-   The classifications are based on the concession type and Anatomical Therapeutic Chemical (ATC) indexes.

We can use `dplr` functions with `tsibble` objects. In this case, we are interested in the `Cost` time series of A10 scripts in millions of dollars.

```{r}
a10 <-
  PBS |>
  filter(ATC2 == "A10") |>
  select(Month, Concession, Type, Cost) |>
  summarize(TotalC = sum(Cost)) |>
  mutate(Cost = TotalC/1e6) # Makes the values more managable
```

`autoplot()` will automatically produce the appropriate plot based on what is passed to it as the first argument.

-   We can also specify the argument that `autoplot()` should use in order to generate the plot (e.g. Cost)

```{r}
autoplot(a10, Cost) + #Specifies that we are plotting Cost
  labs(y = "$ Millions",
       title = " Austrialian diabetes drug sales")
```

-   There is a increasing trend with a strong seasonal pattern that increases in size over time.

-   The sudden drop in the beginning of the year is due to the government subsidization scheme. This incentives patients to stockpile drugs at the end of the year to save money.

Another interesting time series is flights between Australia's two largest cities. Below is a plot of the weekly economy passengers (in thousands) between Melbourne and Sydney.

```{r}
melsyd_economy <-
  ansett |>
  filter(Airports == "MEL-SYD", Class == "Economy") |>
  mutate(Passengers = Passengers/1000)

autoplot(melsyd_economy, Passengers) + 
  labs(title = "Ansett airlines economy class",
       subtitle = "Melbourne-Sydney",
       y = "Passengers (Thousands)")
```

-   There is a large drop in the late 1980s due to pilot strikes where no passengers traveled.

-   There is also a big dip in 1992, where the airline changed economy seats to business class. The level of passengers returned to normal levels when the airline reversed their decision.

-   The dips in number of passengers at the beginning of each year is due to holiday effects.

Below is a plot that verifies the swap between business and economy seats in 1992:

```{r}
ansett |>
  filter(Airports == "MEL-SYD") |>
  mutate(Passengers = Passengers/1000) |>
  autoplot(Passengers) +
  labs(title = "Ansett airlines by class type",
       subtitle = "Melbourne-Sydney",
       y = "Passengers (Thousands)")
```

## Seasonal Plots

Monthly seasonal plot for diabetic drug sales

```{r}
a10 |>
  gg_season(Cost, labels = "both") +
  labs(y = "$ (millions)",
       title = "Seasonal plot: Diabetic drug sales")
```

### Victoria Electricity Demand

The `vec_elec` dataset contains half-hourly electric demand for Victoria Australia.

A seasonal plot with a yearly pattern shows seasonality and that usage remained high in later into Q1 in 2013 than the other years.

```{r}
vic_elec |>
  gg_season(Demand, period = "year") + 
  labs(y = "MWh",
       title = "Victoria Electricity Demand")
```

-   There are so many observations that it is hard to gain much insight from the yearly plot.

#### Weekly demand

Addition trends can be seen when plotting seasonal electricity demand with a weekly pattern

```{r}
vic_elec |>
  gg_season(Demand, period = "week") +
  labs(y = "MWh",
       title = "Victoria energy demand")
```

-   Week days have higher demand than weekend days.

-   Each day has two spikes in demand, likely when people wake up to get ready for work and again when they get home. The spikes are smaller on weekends likely due to less regimented schedules. When people can sleep in, demand is spread out.

#### Daily demand

Daily demand will show seasonality over the hours in a day.

```{r}
vic_elec |>
  gg_season(Demand, period = "day") +
  labs(y = "MWh",
       title = "Victoria electricy demand")
```

-   The three lines that are above normal each year is due to three days a year having 25 hours due to daylight savings time. The three days that have less than 24 hours are lost in the middle.

-   We can see that most people are asleep start using less electricity around 10pm and demand is at it's lowest around 4am.

-   People start waking up around 5am, which increases until about 8am as most people are up for work. There is another equally large spike around 6pm when people come back home from work.

## Seasonal Sub-series Plots

Subseries plots for Australian diabetic drug sales

```{r}
a10 |>
  gg_subseries(Cost) +
  labs(y = "$ Millions",
       title = "Australian diabetic drug sales")
```

-   The blue line indicates the mean for each month.

-   The subseries plot makes it easier to identify changes within a specific season.

The drug sales plot is doesn't reveal any real new information but we can see changes in the Australian holiday tourism (quarterly) dataset. Suppose we want to investigate how quarterly holiday vacation days are spent by state for each quarter.

```{r}
holidays <-
  tourism |>
  filter(Purpose == "Holiday") |>
  group_by(State) |>
  summarize(Trips = sum(Trips))

autoplot(holidays, Trips) + 
  labs(y = "Overnight trips (Thousands)",
       title = "Australian domestic holidays")
```

-   The time plot shows seasonality but the seasonal peaks do not coincide.

The seasonal plot can give us a bit more insight into the seasonality of the data.

```{r}
holidays |>
  gg_season(Trips) +
  facet_wrap(vars(State),
             nrow = 2,
             scales = "free_y") +
  labs(y = "Thousands of trips",
       title = "Austrailian domestic holidays")
```

-   The trends are different in each state.

    -   The Northern Territory and Queensland have their peak travel in Q3, which is the opposite of South Australia, Tasmania, and Victoria.

        -   This is likely due to northern states having the most tourism during their dry season (Q3) and southern ones having their highest levels of tourism during the summer (Q1).

It is easier to see trends in the sub-series plots for each season.

```{r}
holidays |>
  gg_subseries(Trips) +
  labs(y = "Overnight trips (Thousands)",
       title = "Austrailian domestic holidays")
```

## ACF plot with trend and seasonal

Below is a time plot of US retail trade employment

```{r}
retail <-
  us_employment |>
  filter(Title == "Retail Trade",
         year(Month) >= 1980)

retail |> autoplot(Employed)
```

-   Trending upwards with spikes every year around month 12, which is the seasonality

The autocorrelation function for the data above:

```{r}
retail |>
  ACF(Employed, lag_max = 48 ) |>
  autoplot()
```

-   Slight peaks around seasonal multiples (12)

### ACF Plot for Google stock price in 2015

```{r}
google_2015 <-
  gafa_stock |>
  filter(Symbol == "GOOG",
         year(Date) == 2015) |>
  select(Date,
         Close)
head(google_2015)
```

-   The exclamation point in `A tsibble: 6x2 [!]` means that the data is irregularly spaced.

```{r}
google_2015 |> autoplot(Close)
```

-   Overall trend is positive

```{r}
google_2015 |>
  ACF(Close, lag_max = 100) |>
 autoplot()
```

-   Typical shape for stock data or data with trends.

## White noise

```{r}
pigs <- aus_livestock |>
  filter(State == "Victoria",
         Animal == "Pigs",
         year(Month) >= 2014)

pigs |>
  autoplot(Count / 1000) + 
  labs(y = "Thousdans",
       title = "Number of pigs slaughtered in Victoria")
```

-   From the time plot, we can't see any seasonality or trend and therefore may think that this is just white noise.

In order to verify if this is just white noise or not, we can test for it using the ACF plot.

```{r}
pigs |>
  ACF(Count) |>
  autoplot()
```

-   Each one has a 5% chance each lag would be greater than the critical value.

-   Having one outside this range isn't out of the ordinary for 16 lags but 2 may mean there is some seasonality, especially since the first is at lag2 and the seconding being lag12. It's hard to see from the time plot but potentially some slight seasonality.

-   Therefore, this series is <u>not white noise</u>

## Mathematical transformations

Comparing Australian gas production time series with adjusted time series.

```{r}
autoplot(aus_production, Gas) +
  labs(title = "Australian Gas Production Time Plot")
```

Transformed data using `guerrero` to choose lambda:

```{r}
lambda <- aus_production |>
  features(Gas, features = guerrero) |>
  pull(lambda_guerrero)
aus_production |>
  autoplot(box_cox(Gas, lambda)) +
  labs(y = "",
       title = latex2exp::TeX(paste0(
         "Transformed gas production with $\\lambda$ = ",
         round(lambda,2))))
```
